Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-12 paper.tex
Some text on breakpoints

\subsection{Distribution of recombination breakpoint intervals}
\label{sec:breakpoint_intervals}

@@

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358 (
and from from 1522 \&  to C / D in the wide ARG, see Supplementary Tables 1 and 2).

@@

Finally, it is clear from Supplementary Figure 4 that of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron followed by
Delta, are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possibly heterogeneity in
recombination probabilities among lineages.

@@

\subsubsection{Long ARG}
[high level summary of the pango recombinant lineages
and samples in the Long ARG]
The long ARG (subsampled to mid-2022) contains 33
Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
XXX) comprising 749 samples (711 when singleton recombinants are removed). If
the GISAID designations are used, 38 Pango lineages exist in the long ARG,
comprising only 515 samples. In the analyses below, we mainly focus on the
larger number of Nextclade designated samples.

Of the 749 Nextclade recombinant samples, all but 2 are inferred to descend
from a recombination node. These two samples are designated XP, and a likely
explanation for their not have an inferred recombinant origin in the long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing. An improved masking pipeline is an important avenue for
future development (see the Discussion).

@@

The remaining remaining 630 Pango X designated sequences trace back to 50
different recombination nodes. We examine the inferred inheritance structure in
the following subsections.

@@

\subsubsection{Single origin}
We would usually assume that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. Thus, we would also expect our ARG to reflect this evolutionary
history, where we have a all the samples assigned to this recombinant lineage
tracing back to a single recombination node, representing this original
recombination event.

\begin{figure}
\begin{tabularx}{\textwidth}{c}

% \end{footnotesize}
% \\

\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

In Figure~\ref{fig:pango-simple-origin-graph}A we illustrate the case of XA. We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset. Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents. Dotted lines show where this subgraph links to other nodes within the ARG. It is clear from the subgraph that all the XA samples trace to a single originating recombination node.

Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
Pango recombinant lineages that trace back to a single recombination node in
the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
monophyletic clades, in which all of the samples given that Pango designation
descend from a single recombination node.

@@

[Discuss panel B where XAG shares a common recombinant origin with samples from
several other X lineages]

>>>>

\subsection{Distribution of recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before signficant
divergence occurs (during onward transmission), if the two parent sequences
are similar there is no way to distinguish them. Here we define the
interval for a particular breakpoint as the interval over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}, and the left bound can then be derived by
sequence comparison of the parents.

@@

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358
(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
Supplementary Tables 1 and 2).

@@

Finally, it is clear from Supplementary Figure\ref{fig:recomb_mrcas_voc_breakdown} that of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron followed by
Delta, are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possibly heterogeneity in
recombination probabilities among lineages.

@@

\subsubsection{Long ARG}
%% high level summary of the pango recombinant lineages and samples in the Long ARG
The long ARG (subsampled to mid-2022) contains 33
Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
XXX) comprising 749 samples (711 when singleton recombinants are removed). In
contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
comprising only 515 samples. In the analyses below, we mainly focus on the
larger number of Nextclade designated samples.

Of the 749 Nextclade-designated recombinants, we infer that all but two descend
from a recombination node. These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing. An improved masking pipeline is an important avenue for
future development (see the Discussion).

@@

The remaining remaining 630 Pango recombinant samples trace back to 50
different recombination nodes. We examine the inferred inheritance structure in
the following subsections.

@@

\subsubsection{Single origin}
In the absence of genealogical information, a reasonable initial assumption is
that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. We would expect our ARG to reveal evolutionary histories of this
nature, where all the samples assigned to a given recombinant lineage
trace back to one recombination node, representing a single originating
recombination event.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced above all
Nextclade-identified XA samples to the closest other sample strains in our dataset.
Dotted lines show where this subgraph links to other nodes within the ARG, and above
recombination nodes only
ancestral nodes are shown (i.e. cousins of the recombinant parents are omitted).
It is clear from the subgraph that all the XA samples trace to a single originating
recombination node (plotted in black), with a breakpoint position around 22227bp.

%% Discuss panel B where XAG shares a common recombinant origin with samples from several other X lineages

Supplementary Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the
Pango recombinant lineages that trace back to a single recombination node in
the long ARG. [FIXME UPDATE THESE] We can see that XA, XAC, XAE, and XF form
monophyletic clades, in which all of the samples given that Pango designation
descend from a single recombination node.

@@
